---
title: "Single-cell ATAC-seq and multiome analyses"
output:
  html_document: default
  pdf_document: default
date: "`r Sys.Date()`"
---

Refer to Signac Vignettes
Analyzing PBMC scATAC-seq: https://stuartlab.org/signac/articles/pbmc_vignette
Joint RNA and ATAC analysis: https://stuartlab.org/signac/articles/pbmc_multiomic
Motif analysis with Signac: https://stuartlab.org/signac/articles/motif_vignette

## Pacakages Loading

```{r message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Signac)
library(Seurat)
library(EnsDb.Hsapiens.v86)
library(BSgenome.Hsapiens.UCSC.hg38)
library(GenomicRanges)
library(JASPAR2020)
library(TFBSTools)
library(patchwork)
library(ggplot2)
```


## Specify File Paths

```{r}
counts_dir <- "/maps/projects/ralab/data/projects/sc_course/input_data/filtered_feature_bc_matrix/"
frag_path <- "/maps/projects/ralab/data/projects/sc_course/input_data/10k_PBMC_Multiome_nextgem_Chromium_X_atac_fragments.tsv.gz"
metadata_path <- "/maps/projects/ralab/data/projects/sc_course/input_data/10k_PBMC_Multiome_nextgem_Chromium_X_per_barcode_metrics.csv"

seurat_obj_rna <- "/maps/projects/ralab/data/projects/sc_course/output_data/20241015_pbmc_RNA_qc_annotated.rds" # replace it with Robert's scRNA final seurat result

seurat_obj_output <- "/maps/projects/ralab/data/projects/sc_course/output_data/pbmc.ATAC.rds" 
```



## Data Loading

```{r message=FALSE, warning=FALSE}
counts <- Read10X(counts_dir)
metadata <- read.csv(
  file = metadata_path,
  header = TRUE,
  row.names = 1
)
pbmc_rna <- readRDS(seurat_obj_rna)
pbmc_rna <- UpdateSeuratObject(pbmc_rna)
```

## Create Chromatin Assay

```{r}
chrom_assay <- CreateChromatinAssay(
  counts = counts$Peaks,
  sep = c(":", "-"),
  fragments = frag_path,
  min.cells = 10,
  min.features = 200
)

pbmc_atac <- CreateSeuratObject(
  counts = chrom_assay,
  assay = "peaks",
  meta.data = metadata
)

pbmc_atac[['peaks']]
granges(pbmc_atac)
```

## Filter Chromosomes

```{r}
peaks.keep <- seqnames(granges(pbmc_atac)) %in% standardChromosomes(granges(pbmc_atac))
pbmc_atac <- pbmc_atac[as.vector(peaks.keep), ]
```

## Genome Annotation 

```{r warning=FALSE}
annotation <- GetGRangesFromEnsDb(ensdb = EnsDb.Hsapiens.v86)
seqlevels(annotation) <- paste0('chr', seqlevels(annotation))

# Add the gene information to the object
Annotation(pbmc_atac) <- annotation
```

## Quality Metrics and Data Visualization

```{r}
# Compute nucleosome signal score, TSS enrichment score, and fractions
pbmc_atac <- NucleosomeSignal(object = pbmc_atac)
pbmc_atac <- TSSEnrichment(object = pbmc_atac)
pbmc_atac$pct_reads_in_peaks <- pbmc_atac$atac_peak_region_fragments / pbmc_atac$atac_fragments * 100
pbmc_atac$blacklist_ratio <- FractionCountsInRegion(
  object = pbmc_atac, 
  assay = 'peaks',
  regions = blacklist_hg38_unified
)

DensityScatter(pbmc_atac, x = 'nCount_peaks', y = 'TSS.enrichment', log_x = TRUE, quantiles = TRUE)
pbmc_atac$nucleosome_group <- ifelse(pbmc_atac$nucleosome_signal > 2, 'NS > 2', 'NS < 2')
FragmentHistogram(object = pbmc_atac, group.by = 'nucleosome_group')

VlnPlot(
  object = pbmc_atac,
  features = c('nCount_peaks', 'TSS.enrichment', 'blacklist_ratio', 'nucleosome_signal', 'pct_reads_in_peaks'),
  pt.size = 0.1,
  ncol = 5
)
```

## Subsetting, UMAP, and Clustering

```{r}
# Subset data
pbmc_atac <- subset(
  x = pbmc_atac,
  subset = nCount_peaks > 10000 &
    nCount_peaks < 100000 &
    pct_reads_in_peaks > 40 &
    blacklist_ratio < 0.01 &
    nucleosome_signal < 2 &
    TSS.enrichment > 4
)

# Dimensionality reduction and clustering
pbmc_atac <- RunTFIDF(pbmc_atac)
pbmc_atac <- FindTopFeatures(pbmc_atac, min.cutoff = 'q0')
pbmc_atac <- RunSVD(pbmc_atac)
DepthCor(pbmc_atac)

pbmc_atac <- RunUMAP(object = pbmc_atac, reduction = 'lsi', dims = 2:30)
pbmc_atac <- FindNeighbors(object = pbmc_atac, reduction = 'lsi', dims = 2:30)
pbmc_atac <- FindClusters(object = pbmc_atac, verbose = FALSE, algorithm = 3)
DimPlot(object = pbmc_atac, label = TRUE) + NoLegend()
```

## Gene Activity and Feature Plotting

```{r warning=FALSE}
gene.activities <- GeneActivity(pbmc_atac)

# Add the gene activity matrix to the Seurat object as a new assay and normalize it
pbmc_atac[['RNA']] <- CreateAssayObject(counts = gene.activities)
pbmc_atac <- NormalizeData(
  object = pbmc_atac,
  assay = 'RNA',
  normalization.method = 'LogNormalize',
  scale.factor = median(pbmc_atac$nCount_RNA)
)

DefaultAssay(pbmc_atac) <- 'RNA'

FeaturePlot(
  object = pbmc_atac,
  features = c('MS4A1', 'CD3D', 'LEF1', 'NKG7', 'TREM1', 'LYZ'),
  pt.size = 0.1,
  max.cutoff = 'q95',
  ncol = 3
)
```


## Find Transfer Anchors

```{r}
transfer.anchors <- FindTransferAnchors(
  reference = pbmc_rna,
  query = pbmc_atac,
  reduction = 'cca'
)
```

## Transfer Data and Add Metadata

```{r}
predicted.labels <- TransferData(
  anchorset = transfer.anchors,
  refdata = pbmc_rna$predicted.id,
  weight.reduction = pbmc_atac[['lsi']],
  dims = 2:30
)
pbmc_atac <- AddMetaData(object = pbmc_atac, metadata = predicted.labels)
```

## Visualization of Cell Types

```{r}
plot1 <- DimPlot(
  object = pbmc_rna,
  group.by = 'predicted.id',
  label = TRUE,
  repel = TRUE) + NoLegend() + ggtitle('scRNA-seq')

plot2 <- DimPlot(
  object = pbmc_atac,
  group.by = 'predicted.id',
  label = TRUE,
  repel = TRUE) + NoLegend() + ggtitle('scATAC-seq')

plot1 + plot2
```

## Identify Major Predicted IDs

```{r}
predicted_id_counts <- table(pbmc_atac$predicted.id)
major_predicted_ids <- names(predicted_id_counts[predicted_id_counts > 20])
pbmc_atac <- pbmc_atac[, pbmc_atac$predicted.id %in% major_predicted_ids]
Idents(pbmc_atac) <- pbmc_atac$predicted.id
```

## Differential Accessibility Analysis

```{r}
DefaultAssay(pbmc_atac) <- 'peaks'
da_peaks <- FindMarkers(
  object = pbmc_atac,
  ident.1 = "CD4 Naive",
  ident.2 = "CD14+ Monocytes",
  test.use = 'wilcox',
  min.pct = 0.1
)
head(da_peaks)
```

## Plots for Selected Features

```{r}
plot1 <- VlnPlot(
  object = pbmc_atac,
  features = rownames(da_peaks)[1],
  pt.size = 0.1,
  idents = c("CD4 Naive", "CD14+ Monocytes")
)
plot2 <- FeaturePlot(
  object = pbmc_atac,
  features = rownames(da_peaks)[1],
  pt.size = 0.1
)

plot1 | plot2
```

## Find Closest Genes

```{r}
open_cd4naive <- rownames(da_peaks[da_peaks$avg_log2FC > 3, ])
open_cd14mono <- rownames(da_peaks[da_peaks$avg_log2FC < -3, ])


closest_genes_cd4naive <- ClosestFeature(pbmc_atac, regions = open_cd4naive)
closest_genes_cd14mono <- ClosestFeature(pbmc_atac, regions = open_cd14mono)

head(closest_genes_cd4naive)
head(closest_genes_cd14mono)
```

## Coverage Plots

```{r}
regions_highlight <- subsetByOverlaps(StringToGRanges(open_cd4naive), LookupGeneCoords(pbmc_atac, "CD4"))

CoveragePlot(
  object = pbmc_atac,
  region = "CD4",
  region.highlight = regions_highlight,
  extend.upstream = 1000,
  extend.downstream = 1000
)

regions_highlight <- subsetByOverlaps(StringToGRanges(open_cd14mono), LookupGeneCoords(pbmc_atac, "LYZ"))

CoveragePlot(
  object = pbmc_atac,
  region = "LYZ",
  region.highlight = regions_highlight,
  extend.upstream = 1000,
  extend.downstream = 5000
)
```



## Load Motif Data from JASPAR

```{r}
pfm <- getMatrixSet(
  x = JASPAR2020,
  opts = list(collection = "CORE", tax_group = 'vertebrates', all_versions = FALSE)
)
```

## Add Motif Information to the PBMC Object

```{r warning=FALSE}
pbmc_atac <- AddMotifs(
  object = pbmc_atac,
  genome = BSgenome.Hsapiens.UCSC.hg38,
  pfm = pfm
)
```

## Identify Top Differentially Accessible Peaks

```{r}
top.da.peak <- rownames(da_peaks[da_peaks$p_val < 0.005 & da_peaks$pct.1 > 0.2, ])
```

## Find Enriched Motifs in Top Peaks

```{r}
DefaultAssay(pbmc_atac) <- 'peaks'
enriched.motifs <- FindMotifs(
  object = pbmc_atac,
  features = top.da.peak
)
```

## Plot Top Enriched Motifs

```{r}
MotifPlot(
  object = pbmc_atac,
  motifs = head(rownames(enriched.motifs))
)
```

## Run ChromVAR for Motif Variability Analysis

```{r warning=FALSE}
pbmc_atac <- RunChromVAR(object = pbmc_atac, genome = BSgenome.Hsapiens.UCSC.hg38)
DefaultAssay(pbmc_atac) <- 'chromvar'
```

## Visualize the Embedding and Feature Plot

```{r}
p1 <- DimPlot(pbmc_atac, label = TRUE, pt.size = 0.1) + NoLegend()
p2 <- FeaturePlot(
  object = pbmc_atac,
  features = "MA0497.1",
  min.cutoff = 'q10',
  max.cutoff = 'q90',
  pt.size = 0.1
)
p1 + p2
```

## Find Differential Activity of Motifs between Cell Types

```{r}
differential.activity <- FindMarkers(
  object = pbmc_atac,
  ident.1 = "CD4 Naive",
  ident.2 = "CD14+ Monocytes",
  only.pos = TRUE,
  mean.fxn = rowMeans,
  fc.name = "avg_diff"
)
```

## Plot Differential Activity for Top Motifs

```{r}
MotifPlot(
  object = pbmc_atac,
  motifs = head(rownames(differential.activity)),
  assay = 'peaks'
)
```


```{r save-results}
saveRDS(pbmc_atac, seurat_obj_output)
```
```



