---
title: "Novo Nordisk Applied Single Cell Analysis Course October 2024 - scATAC-seq analysis advanced"
output:
  html_document: default
  pdf_document: default
date: "`r Sys.Date()`"
---

###### Modified from Signac Vignettes:
###### Analyzing PBMC scATAC-seq: https://stuartlab.org/signac/articles/pbmc_vignette
###### Motif analysis with Signac: https://stuartlab.org/signac/articles/motif_vignette

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

## [0.] Prerequisites

### [0.0.] Install necessary packages
```{r}
utils::install.packages('ggseqlogo')
```

### [0.1.] Load necessary packages
```{r message=FALSE, warning=FALSE}
base::library(Signac) # scATAC-seq analysis tools
base::library(Seurat) # scRNA-seq analysis tools
base::library(GenomicRanges) # Process ATAC peak regions
base::library(TFBSTools) # Transcription factor binding site analysis
base::library(ggplot2) # Visualization tools
base::library(patchwork) # Combined multiple plots

base::library(EnsDb.Hsapiens.v86) # Human gene annotation database
base::library(BSgenome.Hsapiens.UCSC.hg38) # Human genome sequence database
base::library(JASPAR2020) # Transcription factor motif database
```

### [3.5.] Import previous scATAC seurat object
```{r message=FALSE, warning=FALSE}
pbmc_atac <- base::readRDS(
    file = base::paste0(
      here::here("output_data"),
      "/pbmc_ATAC.rds"
    )
  )
```

---

## [4.] Cell type annotation

### [4.1.] Quantify gene activity
#### Quantify the activity of each gene by summing the fragments intersecting 
#### the gene body and promoter (TSS 2 kb upstream region)
```{r warning=FALSE}
# Generate a gene activity matrix  (It may take ~8 mins)
gene.activities <- Signac::GeneActivity(
    object = pbmc_atac
  )

# Add the gene activity matrix to the Seurat object as a new assay
pbmc_atac[['GeneActivities']] <- Seurat::CreateAssayObject(
    counts = gene.activities
  )

# Normalize gene activity matrix
pbmc_atac <- Seurat::NormalizeData(
    object = pbmc_atac,
    assay = 'GeneActivities',
    normalization.method = 'LogNormalize',
    scale.factor = stats::median(pbmc_atac$nCount_GeneActivities)
  )

# Show gene activity of marker genes on UMAP
Seurat::DefaultAssay(pbmc_atac) <- 'GeneActivities'
```

#### Evaluate by visualization
```{r warning=FALSE}
genes <- base::c(
    'MS4A1', 'CD8A', 'CD4',
    'NKG7', 'TREM1', 'LYZ'
  )

Seurat::FeaturePlot(
    object = pbmc_atac,
    features = genes,
    pt.size = 0.1,
    max.cutoff = 'q95',
    ncol = 2
  )
```

### [4.2.] Transfer cell-type labels from scRNA-seq data
### see [4.11.] in "20241014_scRNAseq_analysis.Rmd"
```{r warning=FALSE}
# Import scRNA-seq seurat object  
pbmc_rna <- base::readRDS(
    file = base::paste0(
      here::here("output_data"),
      "/20241015_pbmc_RNA_qc_annotated.rds"
    )
  )
pbmc_rna <- SeuratObject::UpdateSeuratObject(
    object = pbmc_rna
  )

# Find anchors between scATAC gene activity and scRNA gene expression
transfer.anchors <- Seurat::FindTransferAnchors(
    reference = pbmc_rna,
    query = pbmc_atac,
    reduction = 'cca'
  )

# Transfer cell-type labels from scRNA-seq to scATAC-seq basing on anchors
predicted.labels <- Seurat::TransferData(
    anchorset = transfer.anchors,
    refdata = pbmc_rna$predicted.id,
    weight.reduction = pbmc_atac[['lsi']],
    dims = 2:30 # Exclude the first LSI component
  )
pbmc_atac <- SeuratObject::AddMetaData(
    object = pbmc_atac,
    metadata = predicted.labels
  )

# Since we are using a scRNA+ATAC multiome dataset, we can determine the real 
# correspondence between scRNA and scATAC cells based on their barcodes. 
# Therefore, we transfer the cell types from scRNA-seq to scATAC-seq according 
# to real correspondence to assess the performance of the transfer algorithm.
pbmc_atac@meta.data[,"real.labels"] <- NA
cell.intersected = base::intersect(
    base::colnames(pbmc_atac),
    base::colnames(pbmc_rna)
  )
pbmc_atac@meta.data[cell.intersected,"real.labels"] <- pbmc_rna@meta.data[cell.intersected,"predicted.id"]
pbmc_atac@meta.data[,base::c("predicted.id","real.labels")]
```

### [4.3.] Visualize cell type annotation
```{r}
Seurat::DimPlot(
      object = pbmc_rna,
      group.by = 'predicted.id',
      label = TRUE,
      repel = TRUE
    ) + 
  NoLegend() + 
  ggtitle('scRNA-seq')

Seurat::DimPlot(
      object = pbmc_atac,
      group.by = 'predicted.id',
      label = TRUE,
      repel = TRUE
    ) +
  NoLegend() +
  ggtitle('scATAC-seq, predicted labels')

Seurat::DimPlot(
      object = pbmc_atac,
      group.by = 'real.labels',
      label = TRUE,
      repel = TRUE
    ) +
  NoLegend() +
  ggtitle('scATAC-seq, real labels')
```

### [4.4.] Peak calling for each cell type

#### Cellranger aggregates all cell types together for peak calling, 
#### potentially overlooking cell-type specific peaks in rare cell type, 
#### such as pDCs. This oversight might affect the performance of differential
#### accessibility analysis.  We recommend  calling peaks for each cell type 
#### (see https://stuartlab.org/signac/articles/peak_calling for more 
#### information). However, this process requires installing python package 
#### macs2 and may take hours to compute. Therefore, we will not perform this 
#### step as part of this course.

---

## [5.] Differential accessibility analysis

### [5.1.] Identify differentially accessible peaks between cell types
```{r}
SeuratObject::Idents(pbmc_atac) <- pbmc_atac$predicted.id
SeuratObject::DefaultAssay(pbmc_atac) <- 'peaks'
da_peaks <- Seurat::FindMarkers(
    object = pbmc_atac,
    ident.1 = "CD4 Naive",
    ident.2 = "CD14+ Monocytes",
    test.use = 'wilcox', # Wilcoxon Rank Sum test
    min.pct = 0.1 # peaks detected in 10% cells in either of the two populations
  )
utils::head(da_peaks)
```

### [5.2.] Visualize differential peak
```{r}
# Show peak accessibility using violin plot
Seurat::VlnPlot(
    object = pbmc_atac,
    features = base::rownames(da_peaks)[1], # Specify peak name (example)
    pt.size = 0.1,
    idents = base::c("CD4 Naive", "CD14+ Monocytes")
  )

# Show peak accessibility on UMAP
Seurat::FeaturePlot(
    object = pbmc_atac,
    features = base::rownames(da_peaks)[1],  # Specify peak name (example)
    pt.size = 0.1
  )
```

### [5.3.] Annotate differential peaks to closest genes
```{r}
# Selecte differential peaks as defined by log2(fold change) > 3
open_cd4naive <- da_peaks %>%
  dplyr::filter(avg_log2FC > 3) %>%
  base::rownames()
open_cd14mono <- da_peaks %>%
  dplyr::filter(avg_log2FC < -3) %>% 
  base::rownames()

# Find closest genes
closest_genes_cd4naive <- Signac::ClosestFeature(
    object = pbmc_atac,
    regions = open_cd4naive
  )
closest_genes_cd14mono <- Signac::ClosestFeature(
    object = pbmc_atac,
    regions = open_cd14mono
  )

# Show differential peaks and closest genes
utils::head(closest_genes_cd4naive)
utils::head(closest_genes_cd14mono)
```

### [5.4.] Visualize genomic region
```{r}
# Select highlight regions
regions_highlight <- IRanges::subsetByOverlaps(
    x = Signac::StringToGRanges(open_cd14mono),
    ranges = Signac::LookupGeneCoords(pbmc_atac, "LYZ")
  )

# Plot coverage
Signac::CoveragePlot(
    object = pbmc_atac,
    region = "LYZ",
    region.highlight = regions_highlight,
    extend.upstream = 1000,
    extend.downstream = 5000
  )
```

---

## [6.] Motif analysis

### [6.1.] Load Motif information from JASPAR database
```{r warning=FALSE}
# Extract TF motif from JASPAR database
pfm <- TFBSTools::getMatrixSet(
    x = JASPAR2020,
    opts = base::list(
      collection = "CORE", 
      tax_group = 'vertebrates', 
      all_versions = FALSE
    )
  )

# Add motif information to the Seurat object
pbmc_atac <- Signac::AddMotifs(
    object = pbmc_atac,
    genome = BSgenome.Hsapiens.UCSC.hg38,
    pfm = pfm
  )
```

### [6.2.] Identify enriched motifs in differential peaks
```{r}
# find enriched motifs in differential Peaks
Seurat::DefaultAssay(pbmc_atac) <- 'peaks'
enriched.motifs <- Signac::FindMotifs(
    object = pbmc_atac,
    features = open_cd4naive # Differential peaks identified in [5.3.]
  )

# Plot enriched Motifs
Signac::MotifPlot(
    object = pbmc_atac,
    motifs = utils::head(base::rownames(enriched.motifs))
  )
```

### [6.3.] Computing motif activities with ChromVAR
```{r warning=FALSE}
# Run ChromVAR
pbmc_atac <- Signac::RunChromVAR(
    object = pbmc_atac,
    genome = BSgenome.Hsapiens.UCSC.hg38
  )
Seurat::DefaultAssay(pbmc_atac) <- 'chromvar'

# Visualize motif activity of Mef2c on UMAP
p1 <- Seurat::DimPlot(
      pbmc_atac,
      label = TRUE,
      pt.size = 0.1
    ) +
  NoLegend()
p2 <- Seurat::FeaturePlot(
    object = pbmc_atac,
    features = "MA0497.1", # Mef2c
    min.cutoff = 'q10',
    max.cutoff = 'q90',
    pt.size = 0.1
  ) 
p1 + p2
```

### [6.4.] Identify differentially activity of motifs between cell types
```{r}
diff.activity <- Seurat::FindMarkers(
    object = pbmc_atac,
    ident.1 = "CD4 Naive",
    ident.2 = "CD14+ Monocytes",
    only.pos = TRUE,
    mean.fxn = rowMeans,
    fc.name = "avg_diff"
  )

# Visualize motifs with differential activity
Signac::MotifPlot(
    object = pbmc_atac,
    motifs = utils::head(base::rownames(diff.activity)),
    assay = 'peaks'
  )
```

---

## [7.] scATAC+scRNA multiome analysis: Linking peaks to genes

### [7.1.] Add scRNA-seq data
```{r warning=FALSE}
# Import count matrix
pbmc_count_data <- Seurat::Read10X(
    data.dir = base::paste0(
      here::here("input_data/"),
      "filtered_feature_bc_matrix/"
    )
  )

# Create import count matrix
pbmc_atac[["RNA"]] <- SeuratObject::CreateAssayObject(
    counts = pbmc_count_data$`Gene Expression`[,base::colnames(pbmc_atac)],
    min.cells = 3
  )

# Normalization and PCA
Seurat::DefaultAssay(pbmc_atac) <- "RNA"
pbmc_atac <- Seurat::SCTransform(pbmc_atac)
pbmc_atac <- Seurat::RunPCA(pbmc_atac)
```

### [7.2.] Link peaks to genes
```{r warning=FALSE}
# Compute the GC content for each peak
Seurat::DefaultAssay(pbmc_atac) <- "peaks"
pbmc_atac <- Signac::RegionStats(
    object = pbmc_atac,
    genome = BSgenome.Hsapiens.UCSC.hg38
  )

genes <- base::c("MS4A1") # Insert genes of interest

# Link peaks to genes
pbmc_atac <- Signac::LinkPeaks(
    object = pbmc_atac,
    peak.assay = "peaks",
    expression.assay = "SCT",
    genes.use = genes
  )
```

### [7.3.] Visualize peak-gene links
```{r}
# Set cell types for plotting
idents.plot <- base::c(
    "pre-B cell", "B cell progenitor",
    "CD14+ Monocytes", "CD16+ Monocytes", 
    "CD8 effector", "CD8 Naive"
  )
pbmc_atac <- SortIdents(pbmc_atac)

genes <- base::c("MS4A1") # Insert genes of interest

# Plot peak-gene links for MS4A1
Signac::CoveragePlot(
    object = pbmc_atac,
    region = genes,
    features = genes,
    expression.assay = "SCT",
    idents = idents.plot,
    extend.upstream = 500,
    extend.downstream = 10000
  )
```

---

## [8.] Save Seurat object
```{r}
base::saveRDS(
  object = pbmc_atac,
  file = base::paste0(
      here::here("output_data/"),
      "pbmc_ATAC_advanced.rds"
    )
  )
```

---

## [9.] Additional vignettes for further reading & analysis

### [9.1.] Trajectory analysis of scATAC data
#### https://stuartlab.org/signac/articles/monocle
### [9.2.] Transcription factor footprinting
#### https://stuartlab.org/signac/articles/footprint
### [9.3.] Co-accessible networks
#### https://stuartlab.org/signac/articles/cicero
